NATIONALISTIC BIAS IN SPORT PERFORMANCE EVALUATIONS

### 2020-02-25 second-revision/manuscript.Rmd
Second revision submitted

---
title: "NATIONALIST BIAS IN SPORT PERFORMANCE EVALUATIONS"
subtitle: "An example from the ski jumping World Cup"
runhead: Judging biases in skijumping
author:
- affiliation: Department of Medical Statistics, University of Oslo
  name: Leiv Tore Salte Rønneberg
- affiliation: Department of Sociology and Human Geography, University of Oslo.
  name: Torkild Hovde Lyngstad
- affiliation: European University Institute
  name: Juho Härkönen
date: "14 september 2017"
keywords: nationalist bias, skijumping, judges, fixed effects
abstract: "Illegitimate biases of non-independent judges represent a threat to the legitimacy of international sport competitions. Judges’ nationalities represent one source of bias, as they consciously or subconsciously may prefer athletes from their own nation. Such biases may affect the outcomes of competitions. The literature offers no complete consensus on the magnitude, origin and stability of such biases. In this paper, we shed light on these problems, using international ski jumping as an example. We draw on data from the FIS World Cup competitions in the 2006-2008 and 2015-2016 seasons and estimate a series of fixed-effects models to test hypotheses on nationalist biases. Our results reaffirm suspicions of nationalist bias in major ski jumping competitions, but also show their magnitude is too little to be of major relevance to competition outcomes. The biases vary between nations, but do not change markedly over the study period. "
thanks: "The authors are grateful for financial support from the Småforsk programme at the University of Oslo, and to Marte Knutsson for research assistance in the early phase of establishing the data set.  At publication time, all code and material used to produce this paper (with the exception of the proprietary data) will be available. The authors are also grateful to Matti Nykänen for being an enduring source of inspiration in many aspects of our lives."
output:
  pdf_document:
    keep_tex: yes
    latex_engine: xelatex
    template: /home/torkildl/Dropbox/Setups/thl/rmd-xelatex-template-ms.tex
  html_document: default
bibliography: /home/torkildl/Dropbox/Literature/thl-main-library.bib
anonymous: yes
fontsize: 12pt
geometry: margin=2cm
header-includes:
    - \usepackage{placeins}
    - \usepackage{dcolumn}
    - \usepackage{tabularx}
    - \usepackage{lscape}
    - \usepackage{caption}
fig_caption: false
fig.align: "center"
fig.width: 5
---

@@

Research on biases in basketball illustrates the breadth of the field. In their study of American professional basketball, Price, Remer and Stone [@price_subperfect_2012] found that the referees were biased towards calls that would increase profits for the league. More specifically, in turnovers, the referees tend to favour the home team, the losing team in a particular game, and the losing teams of the play-off series. In turn these biases can lead to increased ticket sales and ad-revenue due to more exciting, closer games; and ultimately even extend the playoff series itself, garnering more revenue. Even though these biases may operate towards profit-maximizing goals, the authors note that they need not result from any conscious behaviour on behalf of the referees.

Two related papers explore a potential racial bias among NBA referees, and its consequences for betting markets [@price_racial_2010, @larsen_racial_2008]. The authors note that there is a significant difference between fouls called when the athlete and referees are of the same race versus when they are not. The difference is large enough to be an important factor in determining game outcomes, making it a viable tactic to pursue in betting markets. Analysing data from two seasons, they find that when the majority of judges are white, betting on the team that has more minutes (expected) played by white players will beat the point spread over 50% of the time.

Nationalistic or cultural bias, where judges favour athletes from their home country or culturally similar ones in a systematic way would be most evident in certain nationalistic competitions such as world cups or Olympics. Evidence of these kinds of biases have been found in many different sports, including rugby [@page_evidence_2010] , muay thai [@myers_evidence_2006], figure skating [@yang_cultural_2006], gymnastics [@callahan_cultural_2016], and surfing [@sampaio_three_2012].

@@

Our reading of the empirical literature on nationalistic and other evaluative biases concludes that there is evidence for such biases in a variety of sports. Given its existence, what are the root causes of such bias and how do they emerge? Theoretically, we can concieve of bias resulting from at least three different mechanisms. We denote these three mechanisms the social psychological mechanism, the cultural legitimacy mechanism and the differential professionalization mechanism, and discuss each of them in turn below.

### The social psychological mechanism: Implicit or explicit nationalist bias

The first social psychological mechanism is the simple folk psychology theory of  judges' nationalistic impulses at the subconscious or even conscious level. They prefer their own to win, and give higher scores to those athletes.  Thus crude form of nationalist bias is similar to what has denoted taste-based discrimination, which refers to individuals’ preferences for certain categories of people and is often used to explain discrimination against women or ethnic minorities in the labour market [@becker1957economics].

Applied to ski jumping, the mechanism refers to judges’ conscious nationalistic impulses that bias their evaluations in favour of their own, and against competitors. A subtler form of this mechanism leads to subconscious bias for jumpers from one’s own country, where own preferences or social expectations lead judges to biased evaluations [@bassett2005rating, @campbell1996nonparametric]. Importantly, this theoretical mechanism does not by itself predict variation in the magnitude of biases across countries. It can be combined with other theoretical ideas to arrive at predictions about any such variation.

### Theory of evaluation work, as applied to nationalist biases in ski jumping

The second way of theorizing bias is to consider the role of judges and their efforts as evaluation work. Evaluative work and the justification for evaluations as a sociological study has been profoundly influences by the works of Boltanski \& Thevenon [@boltanski_justification_1991]. A central tenet in their sociology of evaluations is the room for evaluation.

@@

Applied to ski jumping as a field, judges' practices for scoring represents the room for evaluation. Within this room there are likely coexisting practices that may be correlated with judges' nationalities, giving rise to national variations in judge scores. Some countries, for example, Norway and Finland, enjoy long ski jumping traditions, whereas neighbouring Sweden, an otherwise similar country, does not. Most likely the long traditions also imply that judges from these countries to a larger degree than Swedish judges will define the room for evalution. This became salient when Jan Boklöv, a Swede, pioneered the V-style in ski jumping. His style was not comme-il-faut among judges and competitors in ski jumping at the time, and thus heavily penalized by judges. Eventually he won competitions, due to the technical advantages of the V-style, and other actors started to mimic the style [@muller_performance_2008]. These changes then led to a change in the room for evaluations, where V-style was accepted and it is currently completely dominant in the field [@pfister2007sportification].

Mere national variation in judge scoring practices does not yield nationalistic biases in individual scores. However, when we also consider the fact that competitors are trained and socialized in their respective national ski jumping associations and competitions. This implies that young competitors share the more experienced judges' evaluative standards, and in turn this leads to nationalistic biases in judge scoring.

@@

The third source of national variation in ski jumping evaluations is related to the process of professionalization in sports. An essential ingredient in evaluation are standards. In sports, standards are continuously evolving. The continued professionalization of sports could would suggest a higher degree of standardization and a narrowing of the room for evaluations. Several aspects of contemporary ski jumping suggests that professionalization will limit the magnitude of nationalistic biases.

First, FIS licences judges that takes international, high-level judging assignments. Professionalization of ski jumping entails stricted quality control of judge performances and training. Currently FIS requires national skiing associations to hold at least one national-level judge seminar every year to contribute to educating new judges, and all judges are required to participate in such seminars at least once every two years.

Second, There is also a system in place for auditing judges. Chapter 10 of the FIS Rules documents explains that judges are subject to an evaluation by a FIS Sub-committee working together with a data team, and that their work may result in allotting individual judges with "points" that indicate non-fair scoring practices. If judges accumulate many points, they may not be nominated to higher-level competitions or wholly excluded from judging. Such sanctioned judges may re-qualify for judging assignments after scoring performances from television broadcasts.

@@

The relevance of professionalism for the *national variation* in biases emerges when we combine the idea of evaluative work with the likely different levels of professionalism among the national associations that are responsible for judge training. In countries where ski jumping is a relatively important sport, such as Finland and Norway, the bias should be less due to an assumed stronger professionalization.

Summing up, it seems meaningful to reexamine the existence of nationalist biases, study how these biases vary across the countries from which evaluators are drawn, and study whether nationalist biased have decreased in magnitude over time, due to, for example, the general tendency towards increased professionalization in sports.

In addition to the social psychological, cultural legitimacy, and differential professionalization mechanisms discussed above, there may well be yet other mechanisms that also contribute to nationalistic biases of this kind, for example desires on part of individual judges to grow the sport's influence in their countries and that try to boost their own competitors placements.

@@

Our fourth hypothesis states that nationalistic bias in ski jumping evaluations decrease over time, possibly as a result of the continued professionalization of the sport.

@@

@@

Our analysis proceeds in several steps. First, we estimate nationalistic biases using a sequence of regression models where more complexity is incorporated for each model, each model speaking to one of the first three hypoteses. We estimate the models separately for the two periods we have data for, as there may be have been change over time in the nature of nationalistic biases, the subject of our fourth hypothesis.  In a second step, we estimate home field advantages in both jump performance and stylistic evaluation.

@@

In order to test our third hypothesis, we want to allow for variations in bias between groups of judges from the same nation by interacting the judge fixed-effect in model \#2 with the indicator function for judge and competitor nation correspondence. This implies the followung model:

@@

where $L_j$ is an added judge country fixed-effect. The nation-specific estimates for B provide an assessment of the average bias in scores by judges from different countries.\footnote{In principle, we could obtain individual bias estimates for each judge and then analyze these as samples of judge biases scores grouped by judges' nationalities. This would be very demanding of our data and experiments indicate that estimates would be very imprecise. We thus refrain from doing this, and only allow the bias to vary between the different judge nationalities.} We also rank countries by our selected instrument of ski jumping history, namely the historical record of Olympic ski jumping gold medals in the decades up to and including 2002 (before the seasons we study. This medal variable was constructed by scraping existing public statistics on Olympic medals in ski jumping competitions (men's Large Hill and Normal Hill only), and calculating how many medals each country has collected. The hypothesis is tested by estimating the Spearman rank correlation $\rho$ across the countries. A negative correlation indicates support for the hypothesis.

Our fourth hypothesis regards potential changes in the variation of nationalist bias over time. In order to assess changes in the variation in nationalist bias, we estimate a model akin to model \#2, but with an added interaction with a dummy representing the second season (with the first season as a reference group). The parameter estimate for this interaction effect, $\gamma$ provides a direct test of hypothesis with regard to changes in nationalist bias. The model is written as

\begin{equation}
s_{ijp} = B_{t_1}∙\phi(I=J) + \gamma∙B_{t_2}∙\phi(I=J)+ q_p + \epsilon_{ijp}
\end{equation}

@@

This was the concern of the second hypothesis, were we assumed that the bias level would decrease once we controlled for other, unobserved characteristics of the jump. Model 2b is a response to exactly this challenge, and provides a within-jump estimate of the nationalistic bias. In this model, the coefficient is estimated at 0.065, markedly lower than in the first model. It is statistically significant at the 0.1% level. The magnitude is low, but there is clearly a tendency towards nationalistic bias. In our adjusted second model \#2b, where we also include measures of individual judges' leniency, we obtain a comparable estimate of nationalistic bias (0.075, sig. at 0.1\% level).

The conclusion thus far is that there are nationalistic biases in evaluations made by judges in the FIS World Cup competitions over the last decades. These biases do not disappear once we control rigorously for unobserved characteristics of both the jump itself and the judges. The first two hypotheses are supported by the data analysis. We do not know from these results, whether the remaining nationalistic bias is due to the social psychological, cultural-historical or the differential-professionalization mechanisms. It seems fair to allow all three to play a role.

@@

figure2 <- ggplot(data = nationbias,mapping = aes(x = ocountry, y = estimate)) +
    geom_point(size=3) + 
    geom_pointrange(aes(x=country, y=estimate, ymin = lowci, ymax=hici), fatten = 1.2) +
    #color="grey") +
    #geom_segment(aes(x=country, y=estimate, xend=country, yend=0)) +
    xlab("Nation") + ylab("Strength of bias") +
    theme_minimal() +
    ggtitle("Figure 2. Nation-specific estimates of judges' bias")
figure2

@@

By itself, this piece of evidence is not enough to test hypothesis \#3, as it also suggests several specific mechanisms for the variation in biases, namely cultural-historical legacies and differential professionalization. According to these mechanisms, the bias should be lower in countries where ski jumping has a stronger historical role in sports and where the degree of professionalization is the highest. We are unable to directly adjudicate between these two mechanisms, as there is likely a strong positive correlation between the place of ski jumping in the national sports field and the degree of professionalization.

@@

```{r scrape-gold, message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}
library(rvest)
theurl <- "https://en.wikipedia.org/wiki/List_of_Olympic_medalists_in_ski_jumping"
webpage <- read_html(theurl)

lh_table <- webpage %>% 
    html_nodes("table:nth-child(6)") %>% 
    html_table(header=T)

nh_table <- webpage %>% 
    html_nodes("table:nth-child(11)") %>% 
    html_table(header=T)

@@

The Spearman $\rho$ rank correlation coefficient of nationalist bias and Olympic medals was estimated at ```r round(spearman$estimate,3)``` which thus supports the hypothesis that the magnitude of bias is negatively related to the Olympic medal harvest for the country. If we can assume that historical medal harvest is postively related to the relative place of ski jumping in the national sports field and/or the degree of professionalization of ski jumping, our third hypothesis must be considered supported by the data.

Figure 3 shows the scatterplot between Olympic medals and natonalistic bias estimates. An identical scatterplot that also includes the outlier Slovakia, would yield a slightly less negative relationship.

@@

As can be seen from Table 3, the interaction effect is not measurably different from zero. It is estimated at  ```r round(bias_interact,3)``` which is a miniscule magnitude relative to the main effect of ```r round(bias_main,3)```. All in all, there does not seem to be much change in nationalistic biases over the two time periods our data set covers. The fourth hypothesis must thus be rejected. This implies that if there is an ongoing professionalization of the sport, it has not resulted in lower nationalistic biases among judges.

@@

In sum, we have learned that there are nationalistic biases in ski jumping evaluations. These biases persist when controlling for characteristics of the performance and individual judges.
The bias is less strong among nations with a historical culture of ski jumping. The bias have not declined over a decade of potentially increasing professionalization.

What can we learn from these results in totality? At the very least, we can speculate about the origin of nationalistic biases. The simplistic social pscyhological mechanism is supported by the data, by the mere existence of biases. However, as the social psychology should not be very different across these nations, the variation among nations remains a puzzle. The absence of changes over time in bias strength is weak evidence against the idea that biases are results of "unprofessional practices" and will disappear once professionalization increases. A possibility that remains is that, as biases are stronger among nations where ski jumping is less important in the national sports field, judges (unconsciously) from such nations give higher scores to up the chance of notable, positive results and support their sport on the national scene.

@@

Our estimates of nationalistic biases were also small enough that they would not alter the rankings in any of the contests. A comparison with the within-jump variance indicates that the bias only amounts to around a quarter of a standard deviation (cf. Model 3 in Table 2). In other words, the variation between the judges’ scores that is not due to nationalistic bias is much larger. If this variation is considered to be random deviations from a “true” score, it is clear that competitors should worry more about chance variations in judges’ evaluations (and possible idiosyncratic opinions) than about nationalist biases affecting the outcome of the competition. The ski jumping sport has by implementing several safeguards reduced the importance of nationalistic biases to a minimum.

@@

In this paper, we report an analysis of data from the 2006-2008 and 2015/2016 seasons of the FIS World Cup in Nordic ski jumping.  Our overarching research question is whether there is nationalistic biases in ski jumping. In addition, we also consider whether there is still such bias when influences of the actual performance are ruled out, whether there are national variations in how strong the biases are, due for example to the different positions of ski jumping in national sports fields, and whether they change over time, possibly due to the increased professionalization of the sport.

From our results, we confirm a major tenet of the extant literature, and document nationalistic biases in judges’ evaluations of ski jumping performances. The biases persist even when we use sophisticated fixed effects methods to remove performance-related confounders. There is also quite a bit of variation in how strong biases are between nations, and this variation is seemingly related to the historical importance of ski jumping, with nations with a stronger historical ski jumping culture displaying somewhat lower biases. The mechanism generating this trend is less understood, but may be related to differential professionalization of ski jumping in the participating nations. However, we did not find any evidence that there is a change over the period we study in the magnitude of biases.

@@

Knowledge of the sources of evaluative biases in competitive sports can potentially help us better understand biases and discriminatory practices in other fields, such as student and employee recruitment and the allocation of research grants. Our findings of nationalistic and home-field biases have provided evidence that, in particular when assessed together with other related findings, add to our understanding on bias-generating practices. This knowledge may aid sport organizations, as well as other types of organizations, in their control of similar biases, and produce fairer, performance evaluations in the future.

>>>>

---
title: "NATIONALISTIC BIAS IN SPORT PERFORMANCE EVALUATIONS"
subtitle: "An example from the ski jumping World Cup"
runhead: Judging biases in skijumping
author:
- affiliation: Department of Sociology and Human Geography, University of Oslo.
  name: Torkild Hovde Lyngstad
- affiliation: European University Institute
  name: Juho Härkönen
- affiliation: Department of Medical Statistics, University of Oslo
  name: Leiv Tore Salte Rønneberg
date: "14 september 2017"
keywords: nationalistic bias, skijumping, judges, fixed effects
abstract: "Illegitimate biases of non-independent judges represent a threat to the legitimacy of international sport competitions. Judges’ nationalities represent one source of bias, as they consciously or subconsciously may prefer athletes from their own nation. Such biases may affect the outcomes of competitions. The literature offers no complete consensus on the magnitude, origin and stability of such biases. In this paper, we shed light on these problems, using international ski jumping as an example. We draw on data from the FIS World Cup competitions in the 2006-2008 and 2015-2016 seasons and estimate a series of fixed-effects models to test hypotheses on nationalistic biases. Our results reaffirm suspicions of nationalistic bias in major ski jumping competitions, but also show their magnitude is too little to be of major relevance to competition outcomes. The biases vary between nations, but do not change markedly over the study period."
thanks: "The authors are grateful for financial support from the Småforsk programme at the University of Oslo, and to Marte Knutsson for research assistance in the early phase of establishing the data set.  At publication time, all code and material used to produce this paper (with the exception of the proprietary data) will be available. The authors are also grateful to Matti Nykänen for being an enduring source of inspiration in many aspects of our lives."
output:
  pdf_document:
    keep_tex: yes
    latex_engine: xelatex
    template: /home/torkildl/Dropbox/Setups/thl/rmd-xelatex-template-ms.tex
  html_document: default
bibliography: /home/torkildl/Dropbox/Literature/thl-main-library.bib
anonymous: no
fontsize: 12pt
geometry: margin=2cm
header-includes:
    - \usepackage{placeins}
    - \usepackage{dcolumn}
    - \usepackage{tabularx}
    - \usepackage{lscape}
    - \usepackage{caption}
fig_caption: false
fig.align: "center"
fig.width: 5
---

@@

Research on biases in basketball illustrates the breadth of the field. In their study of American professional basketball, Price, Remer and Stone [-@price_subperfect_2012] found that the referees were biased towards calls that would increase profits for the league. More specifically, in turnovers, the referees tend to favour the home team, the losing team in a particular game, and the losing teams of the play-off series. In turn these biases can lead to increased ticket sales and ad-revenue due to more exciting, closer games; and ultimately even extend the playoff series itself, garnering more revenue. Even though these biases may operate towards profit-maximizing goals, the authors note that they need not result from any conscious behaviour on behalf of the referees.

Two related papers explore a potential racial bias among NBA referees, and its consequences for betting markets [@price_racial_2010; @larsen_racial_2008]. The authors note that there is a significant difference between fouls called when the athlete and referees are of the same race versus when they are not. The difference is large enough to be an important factor in determining game outcomes, making it a viable tactic to pursue in betting markets. Analysing data from two seasons, they find that when the majority of judges are white, betting on the team that has more minutes (expected) played by white players will beat the point spread over 50% of the time.

Nationalistic bias, where judges favour athletes from their home country (or otherwise similar ones in terms of national culture)  in a systematic way would be most evident in certain nationalistic competitions such as world cups or Olympics. Evidence of these kinds of biases have been found in many different sports, including rugby [@page_evidence_2010] , muay thai [@myers_evidence_2006], figure skating [@yang_cultural_2006], gymnastics [@callahan_cultural_2016], and surfing [@sampaio_three_2012].

@@

Our reading of the empirical literature on nationalistic and other evaluative biases concludes that there is evidence for such biases in a variety of sports. Given its existence, what are the root causes of such bias and how do they emerge? Theoretically, we can concieve of bias resulting from at least three different mechanisms. We denote these three mechanisms the social psychological mechanism, the cultural legitimacy mechanism and the differential professionalisation mechanism, and discuss each of them in turn below.

### The social psychological mechanism: Implicit or explicit nationalistic bias

The first social psychological mechanism is the simple folk psychology theory of  judges' nationalistic impulses at the subconscious or even conscious level. They prefer contestants from their own country to win, and give higher scores to those athletes.  This crude form of nationalistic bias is similar to what has been denoted taste-based discrimination, which refers to individuals’ preferences for certain categories of people and is often used to explain discrimination against women or ethnic minorities in the labour market [@becker_economics_1957].

Applied to ski jumping, the mechanism refers to judges’ conscious nationalistic impulses that bias their evaluations in favour of contestants from their own country, and against competitors from other countries. A subtler form of this mechanism leads to subconscious bias for jumpers from one’s own country, where own preferences or social expectations lead judges to biased evaluations [@bassettjr_rating_2005, @campbell_nonparametric_1996]. Importantly, this theoretical mechanism does not by itself predict variation in the magnitude of biases across countries. It can be combined with other theoretical ideas to arrive at predictions about any such variation.

### Theory of evaluation work, as applied to nationalistic biases in ski jumping

The second way of theorizing bias is to consider the role of judges and their efforts as evaluation work. Evaluative work and the justification for evaluations as a sociological study has been profoundly influences by the works of Boltanski \& Thevenon [-@boltanski_justification_1991]. A central tenet in their sociology of evaluations is the room for evaluation.

@@

Applied to ski jumping as a field, judges' practices for scoring represents the room for evaluation. Within this room, there are likely coexisting practices that may be correlated with judges' nationalities, giving rise to national variations in judge scores. Some countries, for example, Norway and Finland, enjoy long ski jumping traditions, whereas neighbouring Sweden, an otherwise similar country, does not. Most likely, their long tradition for ski jumping imply that judges from Norway and Finland to a larger degree than Swedish judges will define the room for evalution. This became salient when Jan Boklöv, a Swede, pioneered the V-style in ski jumping. His style was not comme-il-faut among judges and competitors in ski jumping at the time, and thus heavily penalized by judges. Eventually he won competitions, due to the technical advantages of the V-style, and other actors started to mimic the style [@muller_performance_2008]. These changes then led to a change in the room for evaluations, where V-style was accepted and it is currently completely dominant in the field [@pfister_sportification_2007].

Mere national variation in judge scoring practices does not yield nationalistic biases in individual scores. However, when we also consider the fact that competitors are trained and socialized in their respective national ski jumping associations and competitions, this implies that young competitors share the more experienced judges' evaluative standards, which in turn leads to nationalistic biases in judge scoring.

@@

The third source of national variation in ski jumping evaluations is related to the process of professionalisation in sports. An essential ingredient in evaluation are standards. In sports, standards are continuously evolving. The continued professionalisation of sports would suggest a higher degree of standardization and a narrowing of the room for evaluations. Several aspects of contemporary ski jumping suggest that professionalisation will limit the magnitude of nationalistic biases.

First, FIS licences judges to qualify for taking on international, high-level judging assignments. professionalisation of ski jumping entails stricter quality control of judge performances and training. Currently FIS requires national skiing associations to hold at least one national-level judge seminar every year to contribute to educating new judges, and all judges are required to participate in such seminars at least once every two years [cf. @internationalskifederation_fis_2019].

Second, there is also a system in place for auditing judges. Chapter 10 of the FIS Rules documents explains that judges are subject to an evaluation by a FIS Sub-committee working together with a data team, and that their work may result in allotting individual judges with "points" that indicate non-fair scoring practices. If judges accumulate many points, they may not be nominated to higher-level competitions or wholly excluded from judging. Such sanctioned judges may re-qualify for judging assignments after scoring performances from television broadcasts.

@@

The relevance of professionalism for the *national variation* in biases emerges when we combine the idea of evaluative work with the likely different levels of professionalism among the national associations that are responsible for judge training. In countries where ski jumping is a relatively important sport, such as Finland and Norway, the bias should be less due to an assumed stronger professionalisation.

Summing up, it seems meaningful to reexamine the existence of nationalistic biases, study how these biases vary across the countries from which evaluators are drawn, and study whether nationalistic biased have decreased in magnitude over time, due to, for example, the general tendency towards increased professionalisation in sports.

In addition to the social psychological, cultural legitimacy, and differential professionalisation mechanisms discussed above, there may well be yet other mechanisms that also contribute to nationalistic biases of this kind, for example desires on part of individual judges to grow the sport's influence in their countries and that try to boost their own competitors placements.

@@

Our fourth hypothesis states that nationalistic bias in ski jumping evaluations decrease over time, possibly as a result of the continued professionalisation of the sport.

@@

\newpage

@@

Our analysis proceeds in several steps. First, we estimate nationalistic biases using a sequence of regression models where more complexity is incorporated for each model, each model speaking to one of the first three hypoteses. We estimate the models separately for the two periods we have data for, as there may be have been change over time in the nature of nationalistic biases, the subject of our fourth hypothesis.

@@

In order to test our third hypothesis, we want to allow for variations in bias between groups of judges from the same nation by interacting the judge fixed-effect in model \#2 with the indicator function for judge and competitor nation correspondence. This implies the following model:

@@

where $L_j$ is an added judge country fixed-effect. The nation-specific estimates for B provide an assessment of the average bias in scores by judges from different countries.\footnote{In principle, we could obtain individual bias estimates for each judge and then analyze these as samples of judge biases scores grouped by judges' nationalities. This would be very demanding of our data and experiments indicate that estimates would be very imprecise. We thus refrain from doing this, and only allow the bias to vary between the different judge nationalities.} We also rank countries by our selected instrument of ski jumping history, namely the historical record of Olympic ski jumping gold medals in the decades up to and including 2002 (before the seasons we study). This medal variable was constructed by scraping existing public statistics on Olympic medals in ski jumping competitions (men's Large Hill and Normal Hill only), and calculating how many medals each country has collected. The hypothesis is tested by estimating the Spearman rank correlation $\rho$ across the countries. A negative correlation indicates support for the hypothesis.

Our fourth hypothesis regards potential changes in the variation of nationalistic bias over time. In order to assess changes in the variation in nationalistic bias, we estimate a model akin to model \#2, but with an added interaction with a dummy representing the second season (with the first season as a reference group). The parameter estimate for this interaction effect, $\gamma$ provides a direct test of hypothesis with regard to changes in nationalistic bias. The model is written as

\begin{equation}
s_{ijp} = B∙\phi(I=J) + \gamma∙I(t=2)∙\phi(I=J)+ q_p + \epsilon_{ijp}
\end{equation}

@@

This was the concern of the second hypothesis, were we assumed that the bias level would decrease once we controlled for other, unobserved characteristics of the jump. Model 2a is a response to exactly this challenge, and provides a within-jump estimate of the nationalistic bias. In this model, the coefficient is estimated at 0.065, markedly lower than in the first model. It is statistically significant at the 0.1% level. The magnitude is low, but there is clearly a tendency towards nationalistic bias. In our adjusted second model \#2b, where we also include measures of individual judges' leniency, we obtain a comparable estimate of nationalistic bias (0.075, sig. at 0.1\% level).

The conclusion thus far is that there are nationalistic biases in evaluations made by judges in the FIS World Cup competitions over the last decades. These biases do not disappear once we control rigorously for unobserved characteristics of both the jump itself and the judges. The first two hypotheses are supported by the data analysis. We do not know from these results, whether the remaining nationalistic bias is due to the social psychological, cultural-historical or the differential-professionalisation mechanisms. It seems fair to allow all three to play a role.

@@

figure2 <- ggplot(data = nationbias,mapping = aes(x = ocountry, y = estimate)) +
    geom_point(size=3) + 
    geom_pointrange(aes(x=country, y=estimate, ymin = lowci, ymax=hici), fatten = 1.2) +
    #color="grey") +
    #geom_segment(aes(x=country, y=estimate, xend=country, yend=0)) +
    xlab("Nation") + ylab("Strength of bias") +
    theme_minimal() +
    ggtitle("Figure 2. Nation-specific estimates of judges' bias", subtitle="Bars indicate 95% C.I.")
figure2

@@

By itself, this piece of evidence is not enough to test hypothesis \#3, as it also suggests several specific mechanisms for the variation in biases, namely cultural-historical legacies and differential professionalisation. According to these mechanisms, the bias should be lower in countries where ski jumping has a stronger historical role in sports and where the degree of professionalisation is the highest. We are unable to directly adjudicate between these two mechanisms, as there is likely a strong positive correlation between the place of ski jumping in the national sports field and the degree of professionalisation.

@@

```{r scrape-gold, message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}
library(rvest)
theurl <- here::here("second-revision/List of Olympic medalists in ski jumping - Wikipedia.html")

#https://en.wikipedia.org/wiki/List_of_Olympic_medalists_in_ski_jumping"

webpage <- read_html(theurl)

lh_table <- webpage %>%
    html_nodes("table") %>% 
    .[[1]] %>% 
    html_table(header=T)

nh_table <- webpage %>% 
    html_nodes("table") %>% 
    .[[3]] %>% 
    html_table(header=T)

@@

The Spearman $\rho$ rank correlation coefficient of nationalistic bias and Olympic medals was estimated at ```r round(spearman$estimate,3)``` which thus supports the hypothesis that the magnitude of bias is negatively related to the Olympic medal harvest for the country. If we can assume that historical medal harvest is postively related to the relative place of ski jumping in the national sports field and/or the degree of professionalisation of ski jumping, our third hypothesis must be considered supported by the data.

Figure 3 shows the scatterplot between Olympic medals and nationalistic bias estimates. An identical scatterplot that also includes the outlier Slovakia, would yield a slightly less negative relationship.

@@

As can be seen from Table 3, the interaction effect is not measurably different from zero. It is estimated at  ```r round(bias_interact,3)``` which is a miniscule magnitude relative to the main effect of ```r round(bias_main,3)```. All in all, there does not seem to be much change in nationalistic biases over the two time periods our data set covers. The fourth hypothesis must thus be rejected. This implies that if there is an ongoing professionalisation of the sport, it has not resulted in lower nationalistic biases among judges.

@@

In sum, we have learned that there are nationalistic biases in ski jumping evaluations. These biases persist when controlling for characteristics of the performance and individual judges.
The bias is less strong among nations with a historical culture of ski jumping. The bias have not declined over a decade of potentially increasing professionalisation.

What can we learn from these results in totality? At the very least, we can speculate about the origin of nationalistic biases. The simplistic social pscyhological mechanism is supported by the data, by the mere existence of biases. However, as the social psychology should not be very different across these nations, the variation among nations remains a puzzle. The absence of changes over time in bias strength is weak evidence against the idea that biases are results of "unprofessional practices" and will disappear once professionalisation increases. A possibility that remains is that, as biases are stronger among nations where ski jumping is less important in the national sports field, judges (unconsciously) from such nations give higher scores to up the chance of notable, positive results and support their sport on the national scene.

@@

Our estimates of nationalistic biases were also small enough that they would not alter the rankings in any of the contests. A comparison with the within-jump variance indicates that the bias only amounts to around a quarter of a standard deviation (cf. Model 3 in Table 2). In other words, the variation between the judges’ scores that is not due to nationalistic bias is much larger. If this variation is considered to be random deviations from a “true” score, it is clear that competitors should worry more about chance variations in judges’ evaluations (and possible idiosyncratic opinions) than about nationalistic biases affecting the outcome of the competition. The ski jumping sport has, by implementing several safeguards, reduced the importance of nationalistic biases to a minimum.

@@

In this paper, we report an analysis of data from the 2006-2008 and 2015/2016 seasons of the FIS World Cup in Nordic ski jumping.  Our overarching research question is whether there is nationalistic biases in ski jumping. In addition, we also consider whether there is still such bias when influences of the actual performance are ruled out, whether there are national variations in how strong the biases are, due for example to the different positions of ski jumping in national sports fields, and whether they change over time, possibly due to the increased professionalisation of the sport.

From our results, we confirm a major tenet of the extant literature, and document nationalistic biases in judges’ evaluations of ski jumping performances. The biases persist even when we use sophisticated fixed effects methods to remove performance-related confounders. There is also quite a bit of variation in how strong biases are between nations, and this variation is seemingly related to the historical importance of ski jumping, with nations with a stronger historical ski jumping culture displaying somewhat lower biases. The mechanism generating this trend is less understood, but may be related to differential professionalisation of ski jumping in the participating nations. However, we did not find any evidence that there is a change over the period we study in the magnitude of biases. A promising avenue for future research is to exploit the fairly recent increase in activity among female ski jumpers, which would present a less established and professionalised case and may be held to different standards than men [@hornung_darstellung_2014].

@@

Knowledge of the sources of evaluative biases in competitive sports can potentially help us better understand biases and discriminatory practices in other fields, such as student and employee recruitment and the allocation of research grants. Our findings of nationalistic bias is evidence that, in particular when assessed together with other related findings, add to our understanding on bias-generating practices. This knowledge may aid sport organizations, as well as other types of organizations, in their control of similar biases, and produce fairer, performance evaluations in the future.
